Tools for Identifying Partisan Gerrymandering

### 2019-06-18 Latex/_maintext.tex
File Updates

% =================================================================
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% █▀▀ █▀▀ █▀▀ ▀▀█▀▀ ░▀░ █▀▀█ █▀▀▄
% ▀▀█ █▀▀ █░░ ░░█░░ ▀█▀ █░░█ █░░█
% ▀▀▀ ▀▀▀ ▀▀▀ ░░▀░░ ▀▀▀ ▀▀▀▀ ▀░░▀
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    \section{Introduction}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    \lettrine[lines=3, lraise=0.1, nindent=0em]{\color{lightgray}I}{n} \textit{League of Women Voters v. Pennsylvania}, decided January 22, 2018 (henceforth abbreviated \textit{LWV}), the Pennsylvania Supreme Court offered a new and distinctive test for unconstitutional partisan gerrymandering under which it found the state’s congressional districts to be a ``severe and durable'' partisan gerrymander that violated a long-standing provision of the state’s constitution. The congressional map that was struck down was drawn in 2011 by a Republican-controlled legislature and signed into law by the then Republican governor. It was used in elections from 2012 to 2016. Prior to the \textit{LWV} decision, many academics had identified this map as a partisan gerrymander \citep[see e.g.,][]{Royden2017}, with some even identifying it as the most egregious congressional districting in the U.S. \citep[][ esp. Figure 3]{Wang2016_SLR}.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{See also \citet{Mcgann_et_al_2015_ELJ, McGann_et_al_2016_gerrymandering}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    The Pennsylvania Supreme Court held that ``the fact that the 2011 Plan cannot, as a statistical matter, be a plan directed at complying with traditional redistricting requirements is \underline{sufficient} to establish that it violates the Free and Equal Elections Clause'' [of the Pennsylvania State Constitution] (slip op. at p.128, emphasis added).
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{``Elections shall be free and equal; and no power, civil or military, shall at any time interfere to prevent the free exercise of the right of suffrage.'' Pa. Const. art. I, § 5.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    By \textit{traditional} principles (also sometime called \textit{neutral} principles, or \textit{good government} principles), the Court meant drawing contiguous districts, satisfying the \textit{one person, one vote} standard, avoiding diluting the voting strength of protected racial or ethnic minority groups, minimizing unnecessary political sub-unit splits, and providing reasonably compact districts.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Avoiding fragmentation of communities of interest is also sometimes included in this list.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    However, the Court also considered a variety of other indicators involving the partisan aspects of the plan in coming to the conclusion that the map created a ``severe and durable'' partisan gerrymander.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{The trial court considered a variety of factors offered in expert witness testimony, including evidence about the map‘s political impact. The trial court’s findings of fact reveal that the 2011 map exhibited clearly a multiplicity of indicia of partisan gerrymandering. In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander. The trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered. However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents. Instead, it expressed a pious hope that the U.S. Supreme Court, in its review of the appeals of lower court cases, such as the 2017 case of \href{https://www.brennancenter.org/sites/default/files/legal-work/16-1161_Opinion.pdf}{\textit{Gill v. Whitford 585 U. S. \underline{\hspace{3em}} (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply. The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    Having invalidated the \textit{2011 Enacted} map, the Pennsylvania Supreme Court offered deference to the State to draw a remedy plan. The Pennsylvania legislature failed to enact a lawful remedy plan of its own due mostly to divided government that led to disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature. Had such a plan been passed (and signed), the Court would have reviewed its constitutionality. In the absence of a legislatively enacted remedial plan, the Court invited submissions of plans from the parties to the lawsuit, as well the public at large, to submit remedy plans; and it provided guidance as to the features of the plan in which it was most interested. The Court Order in February 2018 inviting submissions of feasible maps mentioned only good government criteria: minimizing the number of county, township and municipality splits, and maximizing district compactness under several different measures of compactness. It made no mention of incumbent protection, nor voting rights act protections, nor of protecting communities of interests, nor did it mention political outcome projections.
\par
    While a remedy plan was offered by Republican leaders from both chambers (the \textit{Joint legislative} plan) and by the Democratic Governor (\textit{Gov. Wolf}), as well as by others, the Court rejected all such plans. Instead, the Court ordered into place for the 2018 election a map of its own drawn for it by a court-appointed consultant. In a split decision, the Court map was endorsed only by judges with Democratic affiliations, though some of the disagreement on the Court had less to do with the map than with the timing of its implementation.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Pennsylvania Supreme Court Judges are elected or retained to ten-year terms in elections with party labels on the ballot. Judges with Democratic affiliations now constitute the majority of the Court. In 2011 this was not the case.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    The initial ruling that the 2011 plan was unconstitutional was sharply attacked by Republicans as judicial overreach, and the map the court majority adopted was rather viciously attacked by Republicans as simply a pro-Democratic gerrymander under the guise of a court-drawn map.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{On March 20, 2018, Rep. Cris Dush, of Jefferson County introduced a measure that would impeach the four Democratic members of the court that voted to replace the map. His argument is that the members of the court overstepped their rights and violated the Separation of Powers principle of the state. Others saw this as simply sour grapes: \\ \href{https://wapo.st/2LlEVvE}{``Pennsylvania Republicans lost the redistricting battle. Now, they’re declaring war on the courts'', https://wapo.st/2LlEVvE (Christopher Ingraham, February 22, 2018).;} \\ \href{https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html}{``Pa. Republicans are talking about impeaching state Supreme Court justices. Do they have an argument?'', https://www.philly.com/philly/news/politics/will-republicans-impeach-pennsylvania-supreme-court-justices-20180222.html (Andrew Seidman \& Jonathan Lai, February 22, 2018).}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    However, the U.S. Supreme Court has declined reviewing either the state court opinion rejecting the 2011 plan or the state court’s chosen map -- presumably because these decision are based entirely on state law.
\par
    In contrast to the \textit{LWV} finding that the 2011 Pennsylvania congressional map was unconstitutional under state law, a 2017 challenge to the Pennsylvania congressional map under the Elections Clause of the U.S. Constitution was rejected by a three-judge federal court by a vote of 2-1. Because the \textit{LWV} decision mooted the federal litigation, the federal challenge was not appealed to the U.S. Supreme Court. However, a 2018 remand of a Wisconsin partisan gerrymandering case, \textit{Gill v. Whitford} (referred to interchangeably as \textit{Gill}) has, to a limited extent, provided clues about the current [ca. June 2019] state of federal partisan gerrymandering jurisprudence than anything before it.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{The \textit{Gill} remand decision still leaves open the question that has haunted opponents of partisan gerrymandering for the last thirty plus years, namely `When, if ever, will the U.S. Supreme Court actually declare a redistricting map unconstitutional?'.  The potential for such a finding of unconstitutionality is laid out in \textit{Davis v. Bandemer, 478 U.S. 109 (1986)}, which declared partisan gerrymandering justiciable, but under which the Supreme Court has never yet (as of June 2019) struck down a plan as an unconstitutional partisan gerrymander. Despite the 7-2 decision on the Wisconsin remand, which rejected the view that the Court should simply bow out of the process of reviewing maps as possible unconstitutional partisan gerrymanders, the retirement of Justice Kennedy and his replacement by a more conservative justice creates even greater uncertainty about whether, in the lifetime of the present authors, there will ever be a U.S. Supreme Court majority that rejects some challenged district (or plan) as an unconstitutional partisan gerrymander. We will relatively soon have the answer to that question due to the inevitable appeals of partisan gerrymandering cases from North Carolina and Maryland that have already had oral argument on March 26, 2019 before the U.S. Supreme Court (\textit{Rucho, et. al}. v. \textit{Common Cause, et. al} [1:16-CV-1026]) and \textit{Lamone, et. al.} v. \textit{Benisek, et. al.} [1:13-cv-03233-JKB]), with decisions expected in June 2019. In addition, there is a partisan gerrymandering challenge in Michigan, \textit{League of Women Voters of Michigan} v. \textit{Johnson} that will certainly be appealed to the U.S. Supreme Court, as will the next lower court decision in \textit{Gill} itself.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    In that remand decision (\textit{Gill v. Whitford, 585 U.S. \underline{\hspace{3em}} (2018)}), the Supreme Court indicated that any standard for partisan gerrymandering must be a challenge to individual districts which demonstrates that the voters in the particular district have had their votes either ``packed'' or ``cracked.''
\par
    Districts that have been \textit{packed} are those which have had voters from one partisan leaning grouped together so as to waste votes of the group because they win the district by an overwhelming majority, thus preventing those excess votes from being used more efficiently in other districts. Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district. The standard for unconstitutional gerrymandering at the federal level -- if there is ever one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the U.S. Supreme Court appears to be insisting on a district-specific test of packing and cracking. 
\par
    We can identify five key features of the \textit{LWV} ruling: (a) though some districts might be particularly egregious, evidence of unconstitutional gerrymandering can be found in a plan as a whole; (b) evidence of purposeful discrimination was not necessary in order to find unconstitutionality; (c) evidence of discriminatory partisan effects was not necessary to show a violation; (d) evidence that a plan complied with good government criteria is almost certainly not sufficient to rule out a finding of unconstitutionality, even though a finding that a plan failed to comply with such criteria was sufficient to invalidate the plan; and (e) statistical evidence about the likelihood that observed features of a plan could not be explained either by chance or by electoral geography played a pivotal role in the decision. 
\par
    Although only directly applicable in Pennsylvania, \textit{LWV} is, we believe nonetheless, the most important partisan gerrymandering case yet decided in a final and definitive fashion this decade. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Indeed, if the Supreme Court continues to dodge any responsibility for checking egregious partisan gerrymandering, \textit{LWV} may well prove to be the most important partisan gerrymandering decision ever \citep{Grofman_Cervas_2018_ELJ}.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    Because the U.S. Constitution provides for the minimum rights provided to all citizens, state laws and constitutions offer additional protections. Language virtually identical to the provision of the PA state constitution that the Pennsylvania Supreme Court relied on is found in the constitutions of another dozen states, and another thirteen have very similar language \citep{Douglas2014_RightToVote}.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{See also \citet{Elmendorf2018} and further discussion in the conclusions section of this paper.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    That language could be readily adapted to control gerrymandering in those states in much the same way as did the Pennsylvania court. In fact, in its opinion, the Pennsylvania Supreme Court offers what we read as direct encouragement to state courts in other states to follow the lead of Pennsylvania in crafting a way to overturn partisan gerrymandering as a violation of a \underline{state constitution}.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{Footnote 71 of the Opinion (slip op. pp.116-117)}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
\par
    Some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts. Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them. 
\par
    Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard. Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked'' to address what \citet[][see also \citeyear{Grofman_Cervas_2018_ELJ}]{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact. The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map. 
\par
    In the remainder of this essay we will not seek to review in any more detail the various legal and expert witness issues raised in the Pennsylvania litigation itself.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{\citet{Grofman_Cervas_2018_ELJ} provide a review of the various expert witness testimony in the Pennsylvania case.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    Rather, our primary aim is the straightforward one of comparing and contrasting the 2011 and 2018 Pennsylvania congressional maps in terms of a variety of proposed metrics for detecting partisan gerrymandering. In particular, we examine the evidence for the claim that the 2011 map was a pro-Republican gerrymander, and the claim that the 2018 map was a pro-Democratic gerrymander. The former claim is one virtually universally acknowledged \citep{Fang2014, Mcgann_et_al_2015_ELJ, McGann_et_al_2016_gerrymandering, Wang2016_SLR, Wang2016_ELJ}, while the latter claim has been offered not only as partisan griping by Republicans on the losing side of the litigation, but by at least one independent journalist.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{See NYT Upshot: \href{https://nyti.ms/2w8b19k}{``Democrats Didn’t Even Dream of This Pennsylvania Map. How Did It Happen?'', https://nyti.ms/2w8b19k (Nate Cohn, Feb. 21, 2018).;} \\
        \href{https://nyti.ms/2wkpEFT}{``Hundreds of Simulated Maps Show How Well Democrats Fared in Pennsylvania'', https://nyti.ms/2wkpEFT (Nate Cohn, Feb. 26, 2018).}}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    However, the U.S. Supreme Court has declined reviewing either the state court opinion rejecting the 2011 plan or the state court’s chosen map -- presumably because these decision are based entirely on state law.
\par
    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other traditional features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{Joint Legislative} plan offered by Republican legislators from both chambers, and the map proposed by the Democratic governor (\textit{Gov. Wolf}). Here we focus on the question of whether either of these plans was a \textit{stealth gerrymander} done to aid the party drawing the map.
\par
% =================================================================
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% █▀▀ █▀▀ █▀▀ ▀▀█▀▀ ░▀░ █▀▀█ █▀▀▄
% ▀▀█ █▀▀ █░░ ░░█░░ ▀█▀ █░░█ █░░█
% ▀▀▀ ▀▀▀ ▀▀▀ ░░▀░░ ▀▀▀ ▀▀▀▀ ▀░░▀
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    \section{Evaluating the Various Plans -- Methods and Theory}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    To facilitate comparisons to the two most politically important remedy plans offered to the Court, the joint plan proposed by the Republicans leaders of the two legislative chambers, and the Governor’s proposed plan as a Democratic alternative, we look at a number of features in each of the four plans as a whole. We do not, however, discuss district-specific criteria for identifying gerrymandering in these maps, since, although some district-specific evidence was presented about the 2011 map in the trial testimony by an expert who was highly knowledgeable about communities of interest in the state, the Pennsylvania Supreme Court opinion itself offers a \underline{statewide} standard for a finding of unconstitutional partisan gerrymandering. 
   % ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{As noted earlier, given the language in the U.S. Supreme Court’s remand of the \textit{Gill} case, it appears that any federal standard for \underline{partisan gerrymandering} would need to be district specific.}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    Consequently, we limit ourselves to metrics that report results for a jurisdiction as whole.
\par
    All the plans create contiguous districts. Similarly, all four plans appear to satisfy voting rights requirements. Thus, we do not discuss these two ``good government'' criteria further.
\par
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% ▀▄▀▄▀▄ SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- ▄▀▄▀▄▀
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
        \subsection*{Institutional Measures}
% ------ ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% =====================================================================
% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
% ---------------------------------------------------------------------
    \input{Tables/_tab_summaries.tex}
        \begin{center}\textbf{INSERT TABLE \ref{tab:summaries} ABOUT HERE} \end{center}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
\par
    We begin our analyses describing map features distinct from partisan outcomes or proposed measures of partisan asymmetries. Table \ref{tab:summaries} offers a comparison of two key good government features of the map.
%
    \begin{itemize}
        \item the number of county splits, since the Pennsylvania Supreme Court specifically sought to limit the number of counties spanning multiple districts
        \item compactness measures, since compactness is also a criterion specifically referenced by the Pennsylvania Supreme Court
    \end{itemize}
%
\par
% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
            \subsubsection*{County Splits}
% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
    Avoiding county splits is incorporated into the language of many state constitutions \citep{Grofman1985} and can be seen as a way of guarding against gerrymandering by making it more difficult to draw partisan maps. There are a total of 67 counties in PA, and the state is allocated 18 congressional districts. Some county splits are necessitated by the size of the county population relative to the ideal population of a congressional district (705,687). Allegheny and Montgomery must each be split at least twice, while Philadelphia must be split at least three times. If no set of counties exactly equals the ideal congressional population, every districting plan requires a minimum of $n-1$ splits (where $n$ equals the number of congressional districts; assuming that population equality between district must be one person or less). The Pennsylvania Supreme Court found that an unusually and egregiously large number of counties were split in the 2011 congressional map, and they found this fact deeply troubling.
\par
    The data provided in Table \ref{tab:summaries} shows how the high number of county splits in the 2011 map was completely unnecessary. Both the Republican and Democrat proposed remedy plans cut the number of county splits from 41 to 19.
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------
        \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county split. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this ``fracking'', to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find fracking a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool. Instances where counties have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the U.S. District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])).}
% ----------------------------------------------------------------- 
% -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE -- END FOOTNOTE %
% =================================================================
    The Court plan goes even further and gets the number of split counties down to 17.
\par
    Which counties get split and how many ways can impact a plan’s election outcome, which can affect both national politics and a county's relationship with the federal government \citep{Ansolabehere2002a}. That the 2011 map is an eyesore vis-a-vis unnecessary splits of county boundaries is revealed clearly in Figure \ref{fig:maps}.
\par
% =====================================================================
% ▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟
% ---------------------------------------------------------------------
    \input{Figures/_fig_maps.tex}
        \begin{center} INSERT FIGURE \ref{fig:maps} ABOUT HERE \end{center}
% ---------------------------------------------------------------------
% ▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄
% ===================================================================== 
\par
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
            \subsubsection*{Compactness}
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
    The measurement of compactness has long been a central issue in the districting literature \citep[see e.g.,][]{Reock1961, Niemi1990, Polsby1991, Webster2013, Kaufman2018}. Indeed, the original gerrymander was assumed to be evidence of political manipulation because of its irregular shape \citep{Martis2008}. Some minimal level of compactness can make it harder to gerrymander \citep[][cf. \citealt{Webster2013}]{Reock1961, Altman1998}. But compactness is not a magic wand that rules out the possibility of political manipulation. Moreover, as many authors have shown, compactness has multiple dimensions, and these different dimensions need not move in parallel with one another; thus a plan appears compact on one dimension might appear ill-compact on another. The two most common ideas of compactness refer, on the one hand, to how close a legislative district’s boundaries are to its geographic center and, on the other, to how ``regular'' in shape a district appears to be \citep{Niemi1990, Kaufman2018}. In Table \ref{tab:summaries} we report two well-known measures which tap respectively each of these two dimensions. The \textit{Polsby-Popper} measure looks at perimeter irregularity by examining the area of the district compared to that of a circle with similar perimeter, while the \textit{Reock} measure compares the area of a district with that of the district’s circumscribing circle \citep{Reock1961, Polsby1991}. 
% ================================================================= 
% -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE -- FOOTNOTE --  %
% -----------------------------------------------------------------

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% =================================================================
\title{Tools for Identifying Partisan Gerrymandering}
\author{J. Cervas \& B. Grofman}
% 
%
% Jonathan R. Cervas - jcervas@uci.edu - University of California Irvine
% Bernard Grofman - bgrofman@uci.edu - University of California Irvine
%
% 
% First Submitted: Sept 12, 2018
% Revise and Resubmit: January 21, 2019
% Accepted: 
% =================================================================
\documentclass[12pt, a4paper, twoside]{article}
\RequirePackage{cervasanon}
% _________________________________________________________________
%
% _________________________________________________________________
%
% =================================================================
% =================================================================
%                       - Paper Information -
% =================================================================
% =================================================================
%
\begin{document}
    \authorone{Jonathan R. Cervas}
        \institone{University of California Irvine}
    \authortwo{Bernard Grofman}
        \instittwo{University of California Irvine}
    \titlerunning{Tools for Identifying Partisan Gerrymandering}
    \subtitle{With an Application to Congressional Districting in Pennsylvania}
    \status{Revise and Resubmit, Political Geography}
    \data{https://bit.ly/2NaYLOY
    \par
    \centering{\textbf{Key Words}:  gerrymandering, Pennsylvania Supreme Court, redistricting, League of Women Voters, good government}}
    \abstract{In League of Women Voters v. Commonwealth of Pennsylvania (2018) the Pennsylvania Supreme Court struck down as a ``severe and durable” partisan gerrymander the congressional map drawn by Republicans in 2011 and used in elections from 2012-2016. It did so entirely on state law grounds after a three-judge federal court had rejected issuing a preliminary injunction against the plan. After Pennsylvania failed to enact a lawful remedy plan of its own (due to total disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature), the Court then ordered into place for the 2018 election a map of its own drawn for it by a court-appointed consultant. In a split court, the Court map was endorsed only by judges with Democratic affiliations. Here we compare the 2011 and 2018 congressional maps in terms of a variety of proposed metrics for detecting partisan gerrymandering. We also examine the remedy map proposed by a group of Republican legislators and that proposed by the Democratic governor. We conclude that the 2011 map was a blatant and undisguised pro-Republican gerrymander. Moreover, the remedy map proposed by Republican legislators was a covert pro-Republican gerrymander (what we refer to as a ``stealth gerrymander”). The Democratic governor's proposed plan cannot be classified as a pro-Democratic gerrymander and indeed has, if anything, a slight pro-Republican tilt.  The 2018 court-drawn remedial map, by all measures, was not a gerrymander.}

    \thanks{\href{https://www.jonathancervas.com}{Jonathan Cervas} is a Ph.D. candidate in political science at the University of California, Irvine with an interest in elections and skills in geographic information systems. When Grofman was the Special Master in service to the U.S. District Court of Utah, Cervas prepared under his direction remedial maps for the County Commission and the School Board in San Juan County (Navajo Nation et al v. San Juan County et al [12/21/2017]) When Grofman was the Special Master in service to the US District Court for Eastern Virginia, Cervas prepared under his direction remedial maps for Virginia House of Delegates districts (Bethune-Hill et al v. State Board of Election [02/14/2019]). Cervas’s work on this article was supported by the Peltason Chair at UCI and the UCI School of Social Sciences, with supplementary funding from the Peltason Center for the Study of Democracy at UCI.

    \href{http://www.socsci.uci.edu/~bgrofman/}{Bernard Grofman} is the Jack W. Peltason (Bren Foundation Endowed) Chair of Democracy Studies and Distinguished Professor of Political Science at the University of California, Irvine. He has drawn remedial redistricting plans for federal courts and has testified as an expert in numerous redistricting cases over four decades. Most recently. he is the co-author of an Amicus Brief in two 2019 partisan gerrymandering cases that were heard by the U.S. Supreme Court, one from North Carolina and one from Maryland. Grofman’s work on this article was supported by the Peltason Chair at UCI, with supplementary funding from the Peltason Center for the Study of Democracy at UCI. \\

    Questions or comments should be directed to \url{jcervas@uci.edu}. The authors would like to thank \href{https://www.linkedin.com/in/esther-bailey-9289391a/}{Esther Elizabeth Bailey}, \href{https://polisci.wisc.edu/people/faculty/david-canon}{David Canon}, \href{http://www.philly.com/archive/jonathan_lai/}{Jonathan Lai}, \href{https://github.com/modestbayes}{Lingge Li}, \href{mailto:psafarza@uci.edu}{Pooya Safarzadeh}, \href{https://www.faculty.uci.edu/profile.cfm?faculty_id=5443}{Charles Anthony Smith}, \href{https://twitter.com/aruhnv?lang=en}{Aruhn Venkat}, and \href{https://www.dailykos.com/user/Stephen Wolf}{Stephen Wolf} for their helpful suggestions.}
% ================================================================= 
    % ------- TITLE PAGE ------------------------------------------
        \maketitle
        %\input{anon_title.tex}
% =================================================================
\input{_maintext.tex}
% =================================================================
% - Bibliography
%     \input{Tables/_tab_summaries.tex}
% \clearpage
%     \begin{landscape}
%         \input{Tables/_tab_congsum.tex}
%     \end{landscape}
% \clearpage
%     \begin{landscape}
%             \input{Tables/_tab_gerry.tex}
%     \end{landscape}
% \clearpage
%     \begin{landscape}
%         \input{Tables/_tab_prob.tex}
%     \end{landscape}
% % \clearpage
%     \begin{landscape}
%         \input{Tables/_tab_states_compare.tex}
%     \end{landscape}
\clearpage

    % \input{Figures/_maps.tex}

    \nolinenumbers
    \newgeometry{
        top=1in,
        bottom=1in,
        inner=0.85in,
        outer=0.85in,
        ignorehead,
        ignorefoot,
        nomarginpar,
    }

\begin{multicols}{2}
    \footnotesize{\bibliography{library}}   % Bibliography linked to bib.bib
\end{multicols}
% =================================================================
% =================================================================
% APPENDIX
% =================================================================
\setcounter{section}{0} \renewcommand{\thesection}{Appendix \Alph{section}}

    %     \clearpage
    % \input{Tables/a_tab_districtvotes.tex}
    %      \clearpage
    % \input{Tables/a_tab_congress_predict.tex}
    %     \clearpage

\end{document}
